titlD: "Pit-latrine contamination estimates"
output: html_document
datD: '2022-06-24'
---
Develop a 'unique' population raster file for Malawi. Giving the population of Malawi (taken from 2000 population data) combined with proximity to roads and population centres to weight growth towards areas within proximity of settlements and roads

Load packages
```{r}
library(sf)
library(raster)
library(rgdal)
#library(SDMTools)
library(rgeos) 
library(raster)
library(rgdal)
library(rworldmap)
library(maptools)
```

Take shapefile outline of Malawi to clip population
```{r}
download.file("http://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip", "countries.zip")
# Then unzip
unzip("countries.zip")

# Read in the shapefile
world <- readShapeSpatial("ne_10m_admin_0_countries.shp")

plot(world[world$ADMIN=="Malawi",1])

Malawi<- world[world$ADMIN=="Malawi",1]
```

Load Malawi population 2000 data taken from Data Guru. Gives population to 100m resoltion (hence use of this data)
World Pop hub: https://hub.worldpop.org/geodata/summary?id=123
```{r}
pop_2000_file <- raster(paste0('D:/SMaWal/Latrine management/Data/Population/Worldpop/mwi_ppp_2000_UNadj.tif'))
pop_2000<- crop(pop_2000_file,Malawi)
plot(pop_2000)
```

Load data on roads from MASDAP so Malawi specific. http://www.masdap.mw/layers/geonodD:malawi_major_roads
```{r}
roads_Malawi<- readShapeSpatial('D:/SMaWal/Latrine management/Data/Population/Data Guru/malawi_major_roads/malawi_major_roads.shp')
crs(roads_Malawi)<- CRS('+init=EPSG:32736')
```


Generate an empty raster to be used to take the road shape file and convert to a raster of the distance from roads. First outline the size of the raster (taking the extent of the downloaded roads_Malawi dataset), create a raster to these specifications.
```{r}
extent (roads_Malawi)

library(rgeos) 
xmin       <- (477807.6 )
xmax       <- (805224.8) 
ymin       <- (8105221)
ymax       <- (8962213)

distance_raster<- raster(matrix(1:12,953,361), xmn=xmin, xmx=xmax, ymn=ymin, ymx=ymax)
crs(distance_raster)<- CRS('+init=EPSG:32736')
```

Calculate a raster of the distance from each point to the road
```{r}
distance <- gDistance(roads_Malawi, as(distance_raster,"SpatialPoints"), byid=TRUE)
distance_raster[] = apply(distance,1,min)
plot(distance_raster)
dist_roads_original <- distance_raster

dist_roads<- projectRaster(dist_roads_original, crs = CRS('+init=EPSG:4326'))
crs(dist_roads)<- CRS('+init=EPSG:4326')
```

Load data on the fraction of a given cell which is classed as urban
```{r}
urb_frac.stack <-stack(paste('D:/SMaWal/Latrine management/Data/Population/Data Guru/LAND_USE_HARMONIZATION_V1_1248/data/hurtt_urbanfrac_rcp_2_6.tif',sep=''))
urb_frac <- crop(urb_frac.stack[[1]],Malawi,snap='out')
plot(urb_frac)
```

Ensure rasters all equal size
```{r}
pop <- extend(pop_2000,extent(urb_frac))
```

For each grid cell with information on urban fraction, sum together the distance from roads raster, create a centre of gravity raster and all the population raster
```{r}
for(row in 1:nrow(urb_frac)){ 
  for(col in 1:ncol(urb_frac)){ 
     x <-  xFromCol(urb_frac,col) - 0.25 
    y <- yFromRow(urb_frac,row) + 0.25
    ext <- extent(x,x+0.5,y-0.5,y)
    pop_cell <- crop(pop,ext)
    
    if (is.na(maxValue(pop_cell))== TRUE){
      print("NA")
       name<- paste("Pop_cell", row,"_", col, sep = "")
        assign(name, pop_w)}
    
    else if ((is.na(maxValue(pop_cell))== FALSE)){
    ## Distance to road
    dist_roads_cell <- crop(dist_roads,ext)
          dist_roads_cell <- 0.00001 + (max(dist_roads_cell[,])- dist_roads_cell) / (max(dist_roads_cell[,])- min(dist_roads_cell[,])) / 10000
    
          ## Ensure grid cells have same extent and origin
          # template<- projectRaster(pop_cell,dist_roads_cell,method = 'bilinear')
             #projectRaster(from = pop_cell, to= dist_roads_cell, alignOnly=FALSE)
      
          dist_roads_cell<- projectRaster(from = dist_roads_cell, to= pop_cell, alignOnly=FALSE) 
          template<- projectRaster(pop_cell,dist_roads_cell,method = 'bilinear')
          pop_rsmp<- projectRaster(from = pop_cell, to= template)
          
          ##Population centres- rescaled
          cog <- as.numeric(COGravity(pop_rsmp, y = NULL, z = NULL, wt = NULL))
          cog.spdf <- SpatialPointsDataFrame(coords = data.frame(latitude=cog[1],longitude=cog[3],data=1), 
                                             data = data.frame(data=1), proj4string = CRS("+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"))
          cog_dist <- pop_rsmp;  cog_dist[,] <- NA
          cog_dist[cellFromXY(cog_dist, cog.spdf) ] <- 1
          cog_dist <- raster::distance(cog_dist)
          cog_dist <- 0.00001 + (max(cog_dist[,])- cog_dist) / (max(cog_dist[,])- min(cog_dist[,])) / 10000
          
          ## Sum rasters together. Some of the distance to roads raster does not fully complete other grid cells as the roads are too far from the centres. When no distance from roads data is available just use sum of population and centre of population
          
          if (is.na(maxValue(dist_roads_cell))==TRUE){
             pop_w <- pop_rsmp  + cog_dist  
          }
          
             if (is.na(maxValue(dist_roads_cell))==FALSE){
             pop_w <- pop_rsmp + dist_roads_cell + cog_dist 
          }
          
  ## Assign name to the grid cell
          name<- paste("Pop_cell", row,"_", col, sep = "")
        assign(name, pop_w)
        print(name)
  }}}
```

Combine the individual grid cells produced to make large raster of population
```{r}
# Create a list of mask names
mask_names <- paste0("Pop_cell", 1:15, "_", 1:7)

# Use lapply to load the masks and compute the mean mosaic
mosaic_list <- lapply(mask_names, function(mask_name) {
  mask <- get(mask_name)
  mosaic(mask, fun = mean)
})

# Plot each mosaic
lapply(mosaic_list, plot)


pop_unique<- mask(popp, Malawi)

writeRaster(c,'D:/Future population projections Malawi project/unique_population.tif',format='GTiff', overwrite=TRUE)
```

'Unique' starting population raster produced. 
Now to look at historical trends in population growth

Load up required datasets again
```{r}
pop_unique<- raster('D:/Future population projections Malawi project/unique_population.tif')
urb_frac.stack <-stack(paste('D:/SMaWal/Latrine management/Data/Population/Data Guru/LAND_USE_HARMONIZATION_V1_1248/data/hurtt_urbanfrac_rcp_6.tif',sep=''))
urb_frac <- crop(urb_frac.stack[[1]],pop_unique,snap='out')
```

Load historical population information
```{r}
## World bank data for population and IIASA data for urban fraction (for years 2005, 2010 etc extrapolated and using SPPC1 for years 2010-2020)
Historical_population<- read.csv('D:/SMaWal/Latrine management/Data/Population/World Bank Malawi Historical Population.csv')
Historical_population <- Historical_population[,41:61]
Historical_population[2,]<- c(15.176, 15.608,	16.04,	16.472,	16.904,	17.336,	17.8226,	18.3092,	18.7958,	19.2824,	19.769,	20.4986,	20.4986,	20.4986,	20.4986,	23.417,	24.2242,	25.0314,	25.8386,	26.6458,	27.453)
Historical_population[3,]<- (100-Historical_population[2,])
Historical_population[4,]<- (Historical_population[2,]/100)*Historical_population[1,]
Historical_population[5,]<- (Historical_population[3,]/100)*Historical_population[1,]
rownames(Historical_population)<- c("Total population", "Urban Percentage", "Rural percentage", "Urban Population", "Rural population")
```

Starting with the unique population raster, asign population to grid cells
Historic data (no SSP scenarios)
```{r}
### Rank unique population

#Sorting the pixel values within each urban fraction 0.5 degree grid cell and selecting the highest ones as urban until the urban fraction value is fulfilled

population<- pop_unique
scaling_factor<- 1.0

for (i in 1:20){

for(row in 1:((nrow(urb_frac)))){ 
  for(col in 1:((ncol(urb_frac)))){ 
# Locate gridcell, provides coordinates
x <-  xFromCol(urb_frac,col) - 0.25 
y <- yFromRow(urb_frac,row) + 0.25
ext <- extent(x,x+0.5,y-0.5,y)


#print(ext)

pop_cell <- crop(population,ext) # crop population to within urb_frac grid cell

## Apply RCP urban fraction level to rank to define those that count as urban and those that don't
# urb_frac[row, column]. 
RCP_urban_percentage<- ((urb_frac[row,col])/scaling_factor)
#print(RCP_urban_percentage)
RCP_non_urban_percentage<- (1-RCP_urban_percentage)
#print(RCP_non_urban_percentage)

## Which cells have a value exceeding the cut off based on the RCP? 
q<- quantile(pop_cell,RCP_non_urban_percentage)

urban_mask <- reclassify(pop_cell, cbind(-Inf, q, NA))

rural_mask<- reclassify(pop_cell, cbind(q, Inf, NA))

name<- paste("Urban_mask_", row,"_", col, sep = "")
assign(name, urban_mask)

name2<- paste("Rural_mask_", row,"_", col, sep = "")
assign(name2, rural_mask)
  }
  }

# Combine to make urban and rural mask
urban_masks <- lapply(1:15, function(i) {
  lapply(1:7, function(j) {
    get(paste0("Urban_mask_", i, "_", j))
  })
})

rural_masks <- lapply(1:15, function(i) {
  lapply(1:7, function(j) {
    get(paste0("Rural_mask_", i, "_", j))
  })
})

# Create and plot mosaics
mos <- mosaic(urban_masks, fun = mean)
plot(mos)

mos2 <- mosaic(rural_masks, fun = mean)
plot(mos2)



total_urban_population<- cellStats(mos, sum)
urban_mask_percent<- (mos/(total_urban_population))

total_rural_population<- cellStats(mos2, sum)
rural_mask_percent<- (mos2/(total_rural_population))

###
urban_population<- urban_mask_percent*Historical_population[4,(i+1)]
rural_population<- rural_mask_percent*Historical_population[5,(i+1)]

population<- (merge(rural_population,urban_population))

nam <- paste("Malawi_population_200", i, sep = "")
assign(nam, (merge(rural_population,urban_population)))
 
nam <- paste("Rural_population_200", i, sep = "")
assign(nam, ((rural_population)))

nam <- paste("Urban_population_200", i, sep = "")
assign(nam, ((urban_population))) 
 
print(nam)
print(total_urban_population)
print(total_rural_population) 
}
```

```{r}
##Cropping populations by district
population_2001<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Malawi_population_2001_1.0_adjusted_urb_frac.tif'))
urban_population_2001<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Urban_population_2001_1.0_adjusted_urb_frac.tif'))
rural_population_2001<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Rural_population_2001.tif'))
 
# Example Balaka
Balaka_2001_pop<- mask(population_2001,Balaka_sp)
writeRaster(Balaka_2001_pop, 'D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_2001_pop.tif', overwrite=TRUE)
Balaka_urban_2001_pop<- mask(urban_population_2001,Balaka_sp)
writeRaster(Balaka_urban_2001_pop, 'D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_urban_2001_pop.tif', overwrite=TRUE)
Balaka_rural_2001_pop<- mask(rural_population_2001,Balaka_sp)
writeRaster(Balaka_rural_2001_pop, 'D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_rural_2001_pop.tif', overwrite=TRUE)

```
Individual districts percentage of population using latrines rural vs urban for historical years 2010-2022 as data only available from 2010 and 2015 census data. 
Trends in pit latrine usage by looking at change in latrine usage from the 2010 and 2015 census and extrapolating results
```{r}
## Individual district latrine adoption statistics
District_latrines_urban_and_rural<- read.csv('D:/SMaWal/Historic percent pit-latrine usage.csv')
year<- 2010
urban_row<- year-2009
rural_row<- (year+5)-2000
```
```{r}
# For example Balaka 2010
trad_auth<- st_read("D:/SMaWal/QGIS/Malawi/traditional authorities/Malawi_TA_2018.shp")
trad_auth<-st_transform(trad_auth, CRS("+proj=longlat +datum=WGS84 +no_defs"))
trad_auth_single = st_cast(trad_auth,"POLYGON")

Balaka<- trad_auth_single[trad_auth_single$DIST_NAME =="Balaka",]
Balaka_sp<- as_Spatial(Balaka)


## Assume 9.346 users/ latrine urban areas and 12.58 in rural areas
Balaka_population_2010<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_2010_pop.tif'))
Balaka_urban_population_2010<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_urban_2010_pop.tif'))
Balaka_rural_population_2010<- raster(paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_rural_2010_pop.tif'))
Balaka_latrines_urban_population_2010<- (Balaka_urban_population_2010/9.346)*(District_latrines_urban_and_rural[urban_row,]$Balaka/100)
Balaka_latrines_rural_population_2010<- (Balaka_rural_population_2010/12.58)*(District_latrines_urban_and_rural[rural_row,]$Balaka/100)
Balaka_latrines_2010<- merge(Balaka_latrines_urban_population_2010, Balaka_latrines_rural_population_2010)
cellStats(Balaka_latrines_2010,sum)
writeRaster(Balaka_latrines_2010, 'D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_latrines_2010.tif', overwrite= TRUE)


Balaka_latrine_users_urban<- (Balaka_urban_population_2010 )*(District_latrines_urban_and_rural[urban_row,]$Balaka/100)
Balaka_latrine_users_rural<- (Balaka_rural_population_2010 )*(District_latrines_urban_and_rural[rural_row,]$Balaka/100)
Balaka_latrine_users_2010<- merge(Balaka_latrine_users_rural_2010, Balaka_latrine_users_urban_2010)

writeRaster(Balaka_latrine_users_2010, 'D:/Future population projections Malawi project/Population raster files/Historical raster files/Balaka_latrine_users_2010.tif', overwrite= TRUE)
```
### Now looking at future population projections (beyond 2020). Here we look at different scenarios of population growth outlined in SSP scenarios. 
# We start at the first step again. 

# Read in the shapefile,
## Get country outline shapefile (Malawi)
world <- readShapeSpatial("D:/Future population projections Malawi project/Countries zip/ne_10m_admin_0_countries.shp")

plot(world[world$ADMIN=="Malawi",1])

Malawi<- world[world$ADMIN=="Malawi",1]
```

```{r}
# Load urban fraction from Hurtt et al., 2011 data. 
urb_frac.stack <-stack(paste('D:/SMaWal/Latrine management/Data/Population/Data Guru/LAND_USE_HARMONIZATION_V1_1248/data/hurtt_urbanfrac_rcp_2_6.tif',sep=''))
# Crop urban fraction for country case-study
urb_frac <- crop(urb_frac.stack[[1]],Malawi,snap='out')
```

```{r}
### SSP population and urban fraction projections
# Load SSP projection from Riahi et al., (2017) SSP projection database 
SSP_urban_fraction <- read.csv("D:/SMaWal/Latrine management/Data/SSP projections/Malawi SSP projections SSP database.csv")

SSP_population<- read.csv("D:/SMaWal/Latrine management/Data/SSP projections/Malawi population projections under SSP scenarios SSP database.csv")

#Categorise projected populations by SSP scenario

SSP1_pop<- SSP_population[SSP_population$Scenario=="SSP1",]
SSP2_pop<- SSP_population[SSP_population$Scenario=="SSP2",]
SSP3_pop<- SSP_population[SSP_population$Scenario=="SSP3",]
SSP4_pop<- SSP_population[SSP_population$Scenario=="SSP4",]
SSP5_pop<- SSP_population[SSP_population$Scenario=="SSP5",]

## All population projection scenarios are the same for different models
# Use "IIASA GDP" model
SSP1_pop<- SSP1_pop[1,]
SSP2_pop<- SSP2_pop[1,]
SSP3_pop<- SSP3_pop[1,]
SSP4_pop<- SSP4_pop[1,]
SSP5_pop<- SSP5_pop[1,]

##
SSP1_urb<- SSP_urban_fraction[SSP_urban_fraction$Scenario=="SSP1",]
SSP2_urb<- SSP_urban_fraction[SSP_urban_fraction$Scenario=="SSP2",]
SSP3_urb<- SSP_urban_fraction[SSP_urban_fraction$Scenario=="SSP3",]
SSP4_urb<- SSP_urban_fraction[SSP_urban_fraction$Scenario=="SSP4",]
SSP5_urb<- SSP_urban_fraction[SSP_urban_fraction$Scenario=="SSP5",]

###
#For example of population projection SSP5 (can use alternative)

PopulationprojectionSSP5 <- data.frame(matrix(ncol = 19, nrow = 2))
naming<- c("2010", "2015", "2020", "2025", "2030", "2035", "2040", "2045", "2050", "2055", "2060", "2065", "2070", "2075", "2080", "2085", "2090", "2095", "2100")
naming2<- c("Urban", "Rural")
colnames(PopulationprojectionSSP5)<- naming
rownames(PopulationprojectionSSP5)<- naming2

for(b in 1:19){
 PopulationprojectionSSP5[1,b]<- ((SSP5_urb[1,(5+b)])/100)*((SSP5_pop[1,(5+b)])*1000000)
 PopulationprojectionSSP5[2,b]<- (1-((SSP5_urb[1,(5+b)])/100))*((SSP5_pop[1,(5+b)])*1000000)
}


PopulationprojectionSSP5 <- PopulationprojectionSSP5[,3:19]
####
PopulationprojectionSSP5yearly<- matrix(NA,2,81)
colnames(PopulationprojectionSSP5yearly)<- c(2020:2100)

for (j in 1:16) {
  i <- ((j - 1) * 5) + 1
  PopulationprojectionSSP5yearly[1, i:(i + 4)] <- PopulationprojectionSSP5[1, j] + (0:4) * (PopulationprojectionSSP5[1, j + 1] - PopulationprojectionSSP5[1, j]) / 5
  PopulationprojectionSSP5yearly[2, i:(i + 4)] <- PopulationprojectionSSP5[2, j] + (0:4) * (PopulationprojectionSSP5[2, j + 1] - PopulationprojectionSSP5[2, j]) / 5
  print(j)
}

```

For SSP scenario SSP5

For years from 2020-2070 
```{r}
### SSP5
#Sorting the pixel values within each urban fraction 0.5 degree grid cell and selecting the highest ones as urban until the urban fraction value is fulfilled
## If projecting year 2056 (example, take the population raster for the year before, 2055)

Malawi_population_2055<- raster( paste0('D:/Future population projections Malawi project/Population raster files/SSP5/Malawi_population_2055.tif'))

#Malawi_population_2020<- raster( paste0('D:/Future population projections Malawi project/Population raster files/Historical raster files/Malawi_population_2020_no_adjusted_urb_frac.tif'))
population<- Malawi_population_2055

## Set urban smoothing factor
smoothing_factor<- 0.4

for (i in 36:50){

for(row in 1:((nrow(urb_frac)))){ 
  for(col in 1:((ncol(urb_frac)))){ 
# Locate gridcell, provides coordinates
x <-  xFromCol(urb_frac,col) - 0.25 
y <- yFromRow(urb_frac,row) + 0.25
ext <- extent(x,x+0.5,y-0.5,y)

pop_cell <- crop(population,ext) 
# crop population to within urb_frac grid cell
## Within pop_cell rank pixels according to population
## Apply RCP urban fraction level to rank to define those that count as urban and those that don't
# urb_frac[row, column]

RCP_urban_percentage<- ((urb_frac[row,col])/smoothing_factor)
#print(RCP_urban_percentage)
RCP_non_urban_percentage<- (1-RCP_urban_percentage)
#print(RCP_non_urban_percentage)

#print(RCP_urban_percentage)
## Which cells have a value exceeding the cut off based on the RCP? 
q<- quantile(pop_cell,RCP_non_urban_percentage)

urban_mask <- reclassify(pop_cell, cbind(-Inf, q, NA))

rural_mask<- reclassify(pop_cell, cbind(q, Inf, NA))

#print("reclassified")

name<- paste("Urban_mask_", row,"_", col, sep = "")
assign(name, urban_mask)

name2<- paste("Rural_mask_", row,"_", col, sep = "")
assign(name2, rural_mask)
## Need to combine together to produce urban mask and rural mask for whole of Malawi
  }
  }
library(gdalUtils)

# Recombine the tiled urban and rural pieces
urban_masks <- lapply(1:15, function(i) {
  lapply(1:7, function(j) {
    get(paste0("Urban_mask_", i, "_", j))
  })
})

rural_masks <- lapply(1:15, function(i) {
  lapply(1:7, function(j) {
    get(paste0("Rural_mask_", i, "_", j))
  })
})

# Create and plot mosaics
mos <- mosaic(urban_masks, fun = mean)
plot(mos)

mos2 <- mosaic(rural_masks, fun = mean)
plot(mos2)


###########################

total_urban_population<- cellStats(mos, sum)
urban_mask_percent<- (mos/(total_urban_population))

total_rural_population<- cellStats(mos2, sum)
rural_mask_percent<- (mos2/(total_rural_population))

###
# Distribute the population according the the percentage of the population defined in each cell of the urban and rural masks
urban_population<- urban_mask_percent*PopulationprojectionSSP5yearly[1,(i+1)]
rural_population<- rural_mask_percent*PopulationprojectionSSP5yearly[2,(i+1)]

population<- (merge(rural_population,urban_population))

nam <- paste("Malawi_population_20", (i+20), sep = "")
 assign(nam, (merge(rural_population,urban_population)))
 
nam <- paste("Urban_population_20", (i+20), sep = "")
 assign(nam, ((urban_population)))
 
nam <- paste("Rural_population_20", (i+20), sep = "")
 assign(nam, ((rural_population)))
 
print(nam)
print(total_urban_population)
print(total_rural_population) 
}

```

## Now looking at latrine usage
# Scenario A (constant adoption)
```{r}
## Latrine adoption scenario pit latrine usage remains same as was in 2015
District_latrines_rural<- read.csv('D:/SMaWal/UNICEF Groundwater atlas/2015 pit latrine usage rural.csv')
District_latrines_rural<- District_latrines_rural[1,]
District_latrines_urban<- read.csv('D:/SMaWal/UNICEF Groundwater atlas/2015 pit latrine usage urban.csv')
District_latrines_urban<- District_latrines_urban[1,]
```
```{r}
## Example Bakala 2070
Balaka_population_2070<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_2070_pop.tif'))
Balaka_urban_population_2070<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_urban_2070_pop.tif'))
Balaka_rural_population_2070<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_rural_2070_pop.tif'))
Balaka_latrines_urban_population_2070<- (Balaka_urban_population_2070/9.346)*(District_latrines_urban$Balaka/100)
Balaka_latrines_rural_population_2070<- (Balaka_rural_population_2070/12.58)*(District_latrines_rural$Balaka/100)
Balaka_latrines_2070<- merge(Balaka_latrines_urban_population_2070, Balaka_latrines_rural_population_2070)
cellStats(Balaka_latrines_2070,sum)
writeRaster(Balaka_latrines_2070, 'D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_latrines_2070.tif', overwrite= TRUE)


Balaka_latrine_users_urban_population_2070<- (Balaka_urban_population_2070 )*(District_latrines_urban$Balaka/100)
Balaka_latrine_users_rural_population_2070<- (Balaka_rural_population_2070 )*(District_latrines_rural$Balaka/100)
Balaka_latrine_users_2070<- merge(Balaka_latrine_users_urban_population_2070, Balaka_latrine_users_rural_population_2070)

writeRaster(Balaka_latrine_users_2070,'D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_latrine_users_2070.tif', overwrite= TRUE)
```

#Scenarios B or C (just change the CSV file path)
```{r}
#Taking the example Balaka 2055

#load the data file giving estimated pit-latrine usage for each district in 2020 and in 2070
Scenario_B<- read.csv('D:/Future population projections Malawi project/Pit-latrine contamination full dataset/Percent pit-latrine use population/Scenario B  pit-latrine usage by district.csv')
District_latrines_rural<- District_latrines_rural[1,]
# Use the lm() function to create a linear model between 2020 and 2070 estimates
year <- c(2020, 2070)
urban_value <- c(Scenario_B$Balaka[1], Scenario_B$Balaka[2]) 
rural_value <- c(Scenario_B$Balaka[4], Scenario_B$Balaka[5]) 
# Create a data frame for the linear model
urban_estimate <- data.frame(year, urban_value)
rural_estimate <- data.frame(year, rural_value)
# Fit a linear regression model

urban_model <- lm(value ~ year, data = urban_estimate)
rural_model <- lm(value ~ year, data = rural_estimate)
# Specify the year for which you want to estimate the value of pit-latrine usage (e.g., 2050)
new_year <- 2055
new_data <- data.frame(year = new_year)

# Use the model to predict the value for the new year
urban_estimate_value <- predict(model, urban_model = new_data)
rural_estimate_value <- predict(model, rural_model = new_data)

#Input the estimated values into the district population

Balaka_population_2055<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_2055_pop.tif'))
Balaka_urban_population_2055<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_urban_2055_pop.tif'))
Balaka_rural_population_2055<- raster(paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_rural_2055_pop.tif'))

Balaka_latrines_urban_population_2055<- (Balaka_urban_population_2055/9.346)*(urban_estimate_value/100)
Balaka_latrines_rural_population_2055<- (Balaka_rural_population_2055/12.58)*(rural_estimate_value/100)

Balaka_latrines_2055<- merge(Balaka_latrines_urban_population_2055, Balaka_latrines_rural_population_2055)

writeRaster(Balaka_latrines_2055, 'D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_latrines_2055.tif', overwrite= TRUE)


Balaka_latrine_users_urban_population_2055<- (Balaka_urban_population_2055 )*(urban_estimate_value/100)
Balaka_latrine_users_rural_population_2055<- (Balaka_rural_population_2055 )*(rural_estimate_value/100)
Balaka_latrine_users_2055<- merge(Balaka_latrine_users_urban_population_2055, Balaka_latrine_users_rural_population_2055)

writeRaster(Balaka_latrine_users_2055,'D:/Future population projections Malawi project/Population raster files/SSP2/Balaka_latrine_users_2055.tif', overwrite= TRUE)
```
## To look at the whole country, merge individual districts together again
Then break down into individual WRUs
```{r}
# Example WRU 1A
WRU<- st_read("D:/SMaWal/Latrine management/Data/WRU/Files/WRA and WRU/WRU.shp")
WRU<-st_transform(WRU, CRS("+proj=longlat +datum=WGS84 +no_defs"))

WRU_single = st_cast(WRU,"POLYGON")
WRU_1A<- WRU_single[WRU_single$WATER_RESO=="1A",]
WRU_1A_sp<- as_Spatial(WRU_1A[1,])

Malawi_latrine_users_2070<- raster( paste0('D:/Future population projections Malawi project/Population raster files/SSP2/Malawi_whole_latrine_users_population_2070.tif'))
year<- 2070

## Create data frame to input estimates of latrine users per WRU
latrines_users_df <- data.frame(matrix(ncol = 2, nrow = 67))

x <- c( "WRU","2070")
colnames(latrines_users_df)<- x

latrines_users_df[1,]$`2070` <- cellStats((mask(Malawi_latrine_users_2070,WRU_1A_sp)),sum)
```
## Create presence/ absence raster to identify where vulnerable boreholes are
```{r}
# Load the raster package
library(raster)

# Set the extent and resolution of the grid (3-arc seconds)
xmin <- 20
xmax <- 40
ymin <- -17
ymax <- -9.5
res <- 1/120  # 3-arc seconds

# Create an empty raster with the specified extent and resolution
r <- raster(extent(xmin, xmax, ymin, ymax), res = res)

# Load data frame with point coordinates of vulnerable water-points
points <- read.csv('D:/Future population projections Malawi project/Pit-latrine contamination full dataset/Boreholes and latrine colocalisation rasters/vulnerable_water_point_locations.csv')

points <- data.frame(latitude = points$Latitude, longitude = points$Longitude)

x_indices <- floor((points$longitude - xmin) / res)
y_indices <- floor((points$latitude - ymin) / res)


# Create an empty raster with presence/absence values
presence_absence_raster <- raster(extent(xmin, xmax, ymin, ymax), res = res)
presence_absence_raster[] <- 0  # Initialize all cells to 0 (absence)

# Assign presence (1) to grid cells corresponding to the point locations
presence_absence_raster[cbind(x_indices, y_indices)] <- 1

# Plot the presence/absence raster
plot(presence_absence_raster, main = "Presence/Absence Raster")

```

```{r}
##Find cases of latrines where there are boreholes in the same cell
raster_boreholes_and_latrines_2020 <- mask(Malawi_latrines_2020, presence_absence_raster, maskvalue=NA)

## Identify cases over given thresholds, e.g. over 100 latrines for year 2070
Total_raster_boreholes_and_latrines_2070_under_100<-sum(freq(raster_boreholes_and_latrines_2070)[1:95,2])
Total_raster_boreholes_and_latrines_2070<-sum(freq(raster_boreholes_and_latrines_2070)[1:((nrow(freq(raster_boreholes_and_latrines_2070)))-1),2])

raster_boreholes_and_latrines_2070_NAs<- freq(raster_boreholes_and_latrines_2070)[(nrow(freq(raster_boreholes_and_latrines_2070))),2]

Proportion_raster_boreholes_and_latrines_2070_under_100<- Total_raster_boreholes_and_latrines_2070_under_100/Total_raster_boreholes_and_latrines_2070

raster_boreholes_and_latrines_2070_100_and_over_number<- Total_raster_boreholes_and_latrines_2070-Total_raster_boreholes_and_latrines_2070_under_100
```

